\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-17 docs/paper/draft.tex
More fixes
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[7]{r}{.8\columnwidth}
  \begin{minted}{prolog}
db('music.db'). func(f1). func(f2).
trace_sql(p_, I, T) :- trace_sql(p, I, T),
                       graph(p, I, _).
graph(call(f2, call(f1), I), I, O) :-
                       graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\heading{Failure-driven refinement}
When a candidate fails verification, \sys treats this as a feedback signal. Instead of discarding the regeneration attempt wholesale, the system uses the failure to revise or re-execute the synthesis phase.
Some forms are:
\begin{itemize}
  \item \emph{Retrying synthesis}: A fresh LLM call with the same plan but a different generation path.
  \item \emph{Feedback loop}: \sys will provide the previous regeneration attempt 
    as an additional context to the LLM, together with an explanation of the properties that the regeneration violates, prompting it to adapt its output based on prior failures.
    This mode is only activated if the verifier knows that the output will not to taint
    followup regeneration attempts.
  \item \emph{Fallback handling}: If a function repeatedly fails validation, \sys may leave the original version intact or mark it for manual inspection.
\end{itemize}

@@

\sys successfully regenerates all programs in the Rosetta Code benchmark suite,
having verified manually that the output is functionally equivalent
to the original, using both the original and translated reference
implementations as guides.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}).
% \sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses some domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

Here, I outline some possible future directions for \sys and its successors:

@@

\section{Evaluation Samples}
This section provides a selection of samples from the evaluation of \sys.
These instances provide some interesting insights into the capabilities of \sys and its shortcomings.
The samples also provide concrete comparison between \sys and a naive LLM-based approach.

>>>>

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[6]{r}{.8\columnwidth}
  \begin{minted}{prolog}
db('music.db'). func(f1). func(f2).
trace_sql(p_, I, T) :- trace_sql(p, I, T),
                       graph(p, I, _).
graph(call(f2, call(f1), I), I, O) :-
                       graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\heading{Failure-driven refinement}
When a candidate fails verification, \sys treats this as a feedback signal. Instead of discarding the regeneration attempt wholesale, the system uses the failure to revise or re-execute the synthesis phase.
The forms are:
\begin{itemize}
  \item \emph{Retrying synthesis}: A fresh LLM call with the same plan but a different generation path.
  \item \emph{Feedback loop}: \sys will provide the previous regeneration attempt 
    as an additional context to the LLM, together with an explanation of the properties that the regeneration violates.
    A key assumption is that the output will not to taint followup regeneration attempts,
    which should hold based on the synthesis plan, which takes query and plugin constraints into account.
  \item \emph{Fallback handling}: If a function fails validation three times, \sys leaves the original version intact or mark it for manual inspection.
\end{itemize}

@@

\sys successfully regenerates all programs in the Rosetta Code benchmark suite,
having verified manually that the output is functionally equivalent
to the original, using both the original and translated reference
implementations as guides.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}).
% \sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

Here, I outline possible future directions for \sys and its successors:

@@

\section{Evaluation Samples}
This section provides a selection of samples from the evaluation of \sys.
These instances provide interesting insights into the capabilities of \sys and its shortcomings.
The samples also provide concrete comparison between \sys and a naive LLM-based approach.
